My JuliaCon proceeding

### 2023-04-21 paper.tex
Add a sentence more
FILL: TF items need a subsection each, eventually
Write a bit about killing floating point NaNs

% JuliaCon proceedings template
\documentclass{juliacon}
\setcounter{page}{1}

@@

@@

\begin{enumerate}
  \item Monitor the propagation of NaNs through the program
  \item Catch instances of NaN kills
  \item Optionally \emph{inject} NaNs to fuzz the program under scrutiny against NaN-kills
\end{enumerate}

@@

@@

The second thing we evaluated was an N-Body simulation library.
We configured FloatTracker to inject a single NaN to see if we could find any NaN kills.
On some runs we noticed that we would get a kill and the program would warn about a NaN and prepare to exit.
Contrary to the error message's claim, however, the program went into an infinite loop.

>>>>

% JuliaCon proceedings template
\documentclass{juliacon}
\setcounter{page}{1}
\usepackage{amsmath}

@@

The IEEE 754 specification for floating-point numbers \cite{IEEEStandardBinary1985} governs how NaNs are handled as arguments to arithmetic functions.
Most of the time, when any or all of the arguments are NaN, the result is also a NaN.
This is good and desirable, because it means that an arithmetic error---even if buried deep in an expression---will still be observable, allowing a user to find and debug it.

However, there are a few places where NaNs flow into an expression, but not out.
We call these instances ``NaN-kills'', and these can lead to subtle logic bugs.
Consider the following example:

\begin{lstlisting}[language = Julia]
function bad_max(lst)
  max_seen = 0.0
  for x in lst
    if ! (x <= max_seen)
      # swap if new val greater
      max_seen = x
    end
  end
  max_seen
end

function good_max(lst)
  foldl(max, lst)
end
\end{lstlisting}

Both \texttt{bad\_max} and \texttt{good\_max} compute the maximum in a list.
However, the \texttt{<=} operator used in the \texttt{bad\_max} implementation does \emph{not} propagate NaNs, so you might not get the result you expected:

\begin{lstlisting}[language = Julia]
bad_max([1, 5, NaN, 4])     # returns 4
good_max([1, 5, NaN, 4])    # returns NaN
\end{lstlisting}

Not only is the result from \texttt{bad\_max} problematic for obscuring the fact that there was a NaN in the list, the result is also wrong!
The built-in \texttt{max} function correctly propagates NaNs, so we can see from its return that we got a NaN somewhere.
It's not entirely obvious \emph{where} that NaN came from---but more on that later.   % this promise answered in "NaN propagation monitoring"

\subsection{Where do NaNs come from?}

IEEE 754 \cite{IEEEStandardBinary1985} lists the following operations (among a few others) where NaNs result:

\begin{itemize}
\item $\frac{0}{0}$ or $\frac{\infty}{\infty}$
\item $0 * \infty$
\item $x \% y$ when $x = \infty$ or $y = 0$
\item $+\infty + -\infty$ or $\infty - \infty$
\item The square root or logarithm of a negative number
\end{itemize}

% Yadda yadda yadda...

\subsection{NaN kills}

Most mathematical operations preserve NaNs.
However, there are some instances where NaNs can disappear, but not before potentially altering the correctness of the program!

% \begin{itemize}
% \item $x \; \texttt{cmp} \; y \rightarrow false$ where \texttt{cmp} is a binary comparison like \texttt{<}, \texttt{>}, \texttt{<=}, etc. and $x$ or $y$ is NaN.
% \item $1^{NaN} \rightarrow 1$
% \item $NaN^{0} \rightarrow 1$
% \end{itemize}

\begin{align*}
x \; \texttt{cmp} \; y &\rightarrow false &\text{where \texttt{cmp} is \texttt{<}, \texttt{>}, \texttt{<=}, etc.} \\
1^{\text{\texttt{NaN}}} &\rightarrow 1 \\
\text{\texttt{NaN}}^{0} &\rightarrow 1
\end{align*}

% Yadda yadda yadda...

% What else do we need to talk about?

@@

\begin{enumerate}
  \item Catch instances of NaN kills
  \item Monitor the propagation of NaNs through the program
  \item Optionally \emph{inject} NaNs to fuzz the program under scrutiny against NaN-kills
\end{enumerate}

@@

\subsection{Catching NaN kills}

\subsection{NaN propagation monitoring}

Remember in our first example with the \texttt{bad\_max} and \texttt{good\_max} functions how we didn't have a way to tell \emph{where} the NaNs originally came from?
That can be a tedious task.
However, FloatTracker makes this a lot easier by watching the life cycle of a NaN.

% FIXME: talk about CSTGs

\subsection{NaN injection}

@@

The second thing we evaluated was an N-Body simulation library.
We didn't find any sensible configuration of input parameters that lead to NaN kills like in the ShallowWaters library.
We turned to the NaN injection capabilities of FloatTracker to fuzz the library to find any lurking bugs.
% FIXME: add hyper-ref to § 4.1: ShallowWaters

We configured FloatTracker to inject a single NaN to see if we could find any NaN kills.
On some runs we noticed that we would get a kill and the program would warn about a NaN and prepare to exit.
Contrary to the error message's claim, however, the program went into an infinite loop.
